Implementing simplified INLA into `aghq`

### 2022-10-17 src/dev_sinla/sinla.Rmd
Small improvements and table of notation

# Ground truth

@@

Let's have a go at computing Equation \@ref(eq:pxigiveny) so that we can compare to the posterior distribution from `tmbstan`.
We simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
$$
\begin{align}
\tilde p(x_i \, | \, y) &= \tilde p_\text{G}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
&\propto \tilde p_\text{G}(x_i \, | \, \hat \theta, y)
\end{align}
$$
We can evaluate this for $i = 1$ via:

```{r}
i <- 1
mean_i <- as.numeric(mean[i])
var_i <- solve(Q)[ i, i]
```

@@

```{r}
plot2 <- plot +
  stat_function(
    data = data.frame(x = c(0.5, 3.5)), aes(x), fun = dnorm, n = 101, 
    args = list(mean = mean_i, sd = sqrt(var_i)), col = cbpalette[2], size = 1) +
  labs(x = "prev_phi[1]", y = "Posterior") +
  theme_minimal()

@@

@@

Another approach we might take is to calculate the full Laplace approximation
$$
\begin{align*}
  \tilde p(x_i \, | \, \theta, y) &\propto \frac{p(x, \theta, y)}{p_\text{G}(x_{-i} \, | \, x_i, \theta, y)} \Big\rvert_{x_{-i} = \mu_{-i}(x_i, \theta)} (\#eq:pxigiventhetay) \\
  &= \frac{p(y, x_i, \mu_{-i}(x_i, \theta), \theta)}{\det(\frac{1}{2 \pi} Q_{-i}(x_i, \theta))^{1/2}}
\end{align*}
$$
This is not practical in some settings as it involves recomputing the Gaussian approximation $p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ for each value of $(x_i, \theta)$ which can be too computationally expensive.
With of a bit of effort, we can implement this in `TMB` by passing in data where elements $i$ and $-i$ of the latent field are named as such (see Appendix for template).

@@

Now loop over doing the Laplace approximation for every index.
In the call to `MakeADFun` we set `random = "phi_prev_minus_i"` to integrate all of the random effects but that of the particular index.

@@

plot3 <- plot2 +
  geom_line(data = df, aes(x = phi_prev_i, y = post_norm), col = cbpalette[5], size = 1)

@@

First, notational differences between Wood (2019) (former) and here (later) are the following:

* $\beta$ vs. $x$ for random effects
* $\hat \beta$ vs $\mu (\theta)$ for the random effects which maximise the template
* $\tilde \beta$ vs $(x_i, \mu(x_i, \theta))$ for the random effects which maximise the template subject to $x_i$ being fixed
* $H$ vs $Q(\theta)$ for the Hessian / precision matrix of the negative log-likelihood with respect to $x$ at $\mu(\theta)$

>>>>

# Ground truth {#ground-truth}

@@

Now let's have a go at computing Equation \@ref(eq:pxigiveny) so that we can compare to the posterior distribution from `tmbstan` we obtained in Section \@ref(ground-truth).
As described above, we simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
$$
\begin{align}
\tilde p(x_i \, | \, y) &= \tilde p_\text{G}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
&\propto \tilde p_\text{G}(x_i \, | \, \hat \theta, y)
\end{align}
$$
We can evaluate this for $i = 1$ via:

```{r}
i <- 1
mean_i <- as.numeric(mean[i])
var_i <- solve(Q)[i, i]
```

@@

```{r message=FALSE, warning=FALSE}
plot2 <- plot +
  stat_function(
    data = data.frame(x = c(0.5, 3.5)), aes(x, col = "Joint Gaussian marginal"), fun = dnorm, n = 101, 
    args = list(mean = mean_i, sd = sqrt(var_i)), size = 1
  ) +
  labs(x = "prev_phi[1]", y = "Posterior", col = "Approximation") +
  scale_colour_manual(values = cbpalette) +
  theme_minimal()

@@

In this section we will continue to use the approximation \@ref(eq:pthetagiveny) for $\tilde p(\theta \, | \, y)$.

@@

For the approximation to the latent field marginal posteriors, another approach we could take is to calculate the full Laplace approximation
$$
\begin{align*}
  \tilde p(x_i \, | \, \theta, y) &\propto \frac{p(x, \theta, y)}{p_\text{G}(x_{-i} \, | \, x_i, \theta, y)} \Big\rvert_{x_{-i} = \mu_{-i}(x_i, \theta)} (\#eq:pxigiventhetay) \\
  &= \frac{p(y, x_i, \mu_{-i}(x_i, \theta), \theta)}{\det(\frac{1}{2 \pi} Q_{-i}(x_i, \theta))^{1/2}}
\end{align*}
$$
integrating out $x_{-i}$ with a Gaussian approximation.
This is not practical in some settings as it involves recomputing the Gaussian approximation $p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ for each value of $(x_i, \theta)$ which can be too computationally expensive.
We will still go through the steps for implementing it here though for demonstration.
With of a bit of effort, it can be implemented in `TMB` by passing in data where elements $i$ and $-i$ of the latent field are explicitly named (see Appendix for the `TMB` template `model1index.cpp`):

@@

Now we can loop over indices $i$ doing the required Laplace approximations.
In the call to `MakeADFun` we set `random = "phi_prev_minus_i"` to integrate all of the random effects but that of the particular index.

@@

plot3 <- plot2 +
  geom_line(data = df, aes(x = phi_prev_i, y = post, col = "Laplace"), size = 1)

@@

| Description | Here | Wood (2019) notation | Notes |
|-------------|---------------|----------------------|----------|
| Random effects | $x$ | $\beta$ | Stringer (2021) use $w$
| Random effects which maximise (the / some) objective for given $\theta$ | $\mu (\theta)$ | $\hat \beta$ | Wood omits dependence on $\theta$ |
| Random effects which maximise (the / some) objective for given $\theta$, subject to $x_i$ being fixed | $(x_i, \mu(x_i, \theta))$ | $\tilde \beta$ | We have $\mu(x_i, \theta))$ having dimension $n - 1$ |
| Hessian / precision matrix of the negative log-likelihood with respect to $x$ at $\mu(\theta)$ | $Q(\theta)$ | $H$ | Unsure if it's preferable to use Hessian or precision notation here |
